---
title: "Somalia Hunger Visualizations"
output: html_document
---
```{r setup, echo = FALSE, message = FALSE}
knitr::opts_knit$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(lubridate)
library(trackr)
library(ggrepel)
```


```{r benchmarks, echo = FALSE}
benchmarks = c("SOM", "SSF", "LIC", "WLD", "FCS")
classifications <- read_excel("Data/country_classifications.xlsx")
low_inc_codes <- classifications |>
  filter(`Income group` == "Low income") |>
  pull(Code)
ssa_codes <- classifications |>
  filter(Region == "Sub-Saharan Africa") |>
  pull(Code)
fcs_codes <- classifications |>
  filter(Economy %in% c("Afghanistan",
                        "Burkina Faso",
                        "Cameroon",
                        "Central African Republic",
                        "Congo, Dem. Rep.",
                        "Ethiopia",
                        "Haiti",
                        "Iraq",
                        "Lebanon",
                        "Mali",
                        "Mozambique",
                        "Myanmar",
                        "Niger",
                        "Nigeria",
                        "Somalia, Fed. Rep.",
                        "South Sudan",
                        "Sudan",
                        "Syrian Arab Republic",
                        "Ukraine",
                        "West Bank and Gaza",
                        "Yemen, Rep.",
                        "Burundi",
                        "Chad",
                        "Comoros",
                        "Congo, Rep.",
                        "Eritrea",
                        "Guinea-Bissau",
                        "Kiribati",
                        "Libya",
                        "Marshall Islands",
                        "Micronesia, Fed. Sts.",
                        "Papua New Guinea",
                        "São Tomé and Príncipe",
                        "Solomon Islands",
                        "Timor-Leste",
                        "Tuvalu",
                        "Venezuela, RB",
                        "Zimbabwe"
                        )) |>
  pull(Code)
country_pops <- wbstats::wb_data(indicator = "SP.POP.TOTL", lang = "en", country = "countries_only") |>
  select(iso3c, SP.POP.TOTL, date) |>
  rename("code" = iso3c, "population" = SP.POP.TOTL, "year" = date)
```

---

## Prevalence of Undernourishment

```{r, echo = FALSE}
## Prevalence of undernourishment (% of population); ind = SN.ITK.DEFC.ZS
pou <- read.csv("Data/API_SN.ITK.DEFC.ZS_DS2_en_csv_v2_331187.csv", skip = 4) |>
  select(c(Country.Name, Country.Code, "X2001":last_col(offset = 3)))
colnames(pou) <- gsub("X", "", colnames(pou))
pou <- pou |>
  pivot_longer(cols = "2001":last_col(),
               names_to = "year",
               values_to = "value")
```



```{r, echo = FALSE, warning = FALSE}
pou_plot <- pou |>
  filter(Country.Code %in% benchmarks) |>
  mutate(label = if_else(year == max(year), as.character(Country.Name), NA_character_),
         highlight = (Country.Code == "SOM")) |>
  ggplot(aes(x = year, y = value, group = Country.Name, color = highlight)) +
  theme_minimal() +
  geom_line() +
  scale_x_discrete(breaks = seq(2001, 2023, by = 5)) +
  theme(
    legend.position = "none") +
  labs(title = "Somalia has Historically been the Most Undernourished Nation in the World",
       subtitle = "Prevalence of Undernourishment, % of Population",
       x = "",
       y = "") +
  scale_color_manual(values = c("#8A969F", "#0071BC")) +
  geom_vline(xintercept = 23, linetype = "dashed", color = "black") +
  annotate("text",
           x = 20.25, 
           y = 68,
           label = "2023: Data ends",
           fontface = "italic",
           size = 3.5) + 
  geom_label_repel(aes(label = label),
                  nudge_x = 7.5,
                  direction = "y",
                  hjust = 0.5,
                  segment.color = NA,
                  na.rm = TRUE)  +
  theme(plot.subtitle = element_text(face = "italic"))
# ggsave("Visualizations/Pct_Undernourishment.png")
pou_plot
```

Somalia's undernourishment rate compared to global and regional averages, 2001–2023. During this time period, it has consistently been the country with the highest level of undernourishment.

```{r, echo = FALSE, warning = FALSE}
## Line chart with FCS Countries

pou_fcs_plot <- pou |>
  filter(Country.Code %in% fcs_codes) |>
  mutate(label = if_else(year == max(year) & Country.Code == "SOM", as.character(Country.Name), NA_character_),
         highlight = (Country.Code == "SOM")) |>
  ggplot(aes(x = year, y = value, group = Country.Name, color = highlight)) +
  theme_minimal() +
  geom_line() +
  scale_x_discrete(breaks = seq(2001, 2023, by = 5)) +
  labs(title = "Prevalence of Undernourishment, % of Population",
       x = "",
       y = "",
       color = "Country") +
  scale_color_manual(values = c("FALSE" = "#8A969F", "TRUE" = "#0071BC"),
                     labels = c("FALSE" = "FCS Countries", "TRUE" = "Somalia")) + 
  geom_vline(xintercept = 23, linetype = "dashed", color = "black") +
  annotate("text",
           x = 20.25, 
           y = 68,
           label = "2023: Data ends",
           fontface = "italic",
           size = 3.5)

pou_fcs_plot
```

Somalia's undernourishment rate plotted against individual fragile and conflict-affected states, 2001–2023.


```{r, echo = FALSE, warning = FALSE}
pou_arrow <- pou |>
  filter(Country.Code %in% fcs_codes) |>
  rename("country" = Country.Name, "date" = year) |>
  filter(date == 2023 | date == 2015) |>
  mutate(value = value / 100) |>
  pivot_wider(names_from = date, values_from = value, names_prefix = "pou_") |>
  filter(!is.na(pou_2015) | !is.na(pou_2023)) |>
  mutate(highlight = (country == "Somalia, Fed. Rep."))

pou_arrow_plot1 <- pou_arrow |>
  ggplot() +
  geom_segment(aes(x = reorder(country, -pou_2023), xend = reorder(country, pou_2023),
                   y = pou_2015, yend = pou_2023, color = highlight),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed")) +
  geom_point(aes(x = reorder(country, -pou_2023), y = pou_2015, color = highlight), size = 1) +
  scale_color_manual(
    values = c("TRUE" = "#0071BC", "FALSE" = "#8A969F"),
    labels = c("TRUE" = "Somalia", "FALSE" = "FCS Countries"),
    name = NULL
  ) +
  labs(x = "",
       y = "",
       title = "Prevalence of Undernourishment Change from 2015-2023",
       subtitle = "% of Population, Sorted by End Year Value") +
  theme_minimal() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent)
pou_arrow_plot1
```




```{r, echo = FALSE, warning = FALSE}
pou_arrow_plot2 <- pou_arrow |>
  ggplot() +
  geom_segment(aes(x = reorder(country, -pou_2015), xend = reorder(country, pou_2015),
                   y = pou_2015, yend = pou_2023,
                   color = highlight),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed")) +
  geom_point(aes(x = reorder(country, -pou_2015), y = pou_2015, color = highlight), size = 1) +
  scale_color_manual(
    values = c("TRUE" = "#0071BC", "FALSE" = "#8A969F"),
    labels = c("TRUE" = "Somalia", "FALSE" = "FCS Countries"),
    name = NULL
  ) +
  labs(x = "",
       y = "",
       title = "Prevalence of Undernourishment Change from 2015-2023",
       subtitle = "% of Population, Sorted by Start Year Value") +
  theme_minimal() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent)
pou_arrow_plot2

```

Change in undernourishment prevalence across FCS countries from 2015 to 2023, ranked by 2023 rate and by 2015 rate. Although Somalia has improved a great deal in recent years, it still lags the rest of the world.

---

## FIES Data

```{r, echo = FALSE}
## Prevalence of moderate or severe food insecurity in the population (%); ind = SN.ITK.MSFI.ZS
insecurity_ms<- read.csv("Data/API_SN.ITK.MSFI.ZS_DS2_en_csv_v2_7955.csv", skip = 4) |>
  select(c(Country.Name, Country.Code, "X2021")) |>
  rename(pct = X2021)

## Creating Makeshift Benchmarks
### world average
wld_avg_ms <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_ms, by = c("code" = "Country.Code")) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "World", Country.Code = "WLD")
### ssa average
ssa_avg_ms <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_ms, by = c("code" = "Country.Code")) |>
  filter(code %in% ssa_codes) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "Sub-Saharan Africa", Country.Code = "SSF")
### low income average
lic_avg_ms <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_ms, by = c("code" = "Country.Code")) |>
  filter(code %in% low_inc_codes) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "Low income", Country.Code = "LIC")
### fcs average
fcs_avg_ms <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_ms, by = c("code" = "Country.Code")) |>
  filter(code %in% fcs_codes) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "FCS Countries", Country.Code = "FCS")
### Joining to main dataset and filtering
insecurity_ms <- insecurity_ms |>
  filter(Country.Code == "SOM") |>
  bind_rows( wld_avg_ms, ssa_avg_ms, lic_avg_ms, fcs_avg_ms)
  


## Prevalence of severe food insecurity in the population (%); ind = SN.ITK.SVFI.ZS
insecurity_s <- read.csv("Data/API_SN.ITK.SVFI.ZS_DS2_en_csv_v2_331188.csv", skip = 4) |>
  select(c(Country.Name, Country.Code, "X2021")) |>
  rename(pct = X2021)

## Creating Makeshift Benchmarks
### world average
wld_avg_s <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_s, by = c("code" = "Country.Code")) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "World", Country.Code = "WLD")
### ssa average
ssa_avg_s <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_s, by = c("code" = "Country.Code")) |>
  filter(code %in% ssa_codes) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "Sub-Saharan Africa", Country.Code = "SSF")
### low income average
lic_avg_s <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_s, by = c("code" = "Country.Code")) |>
  filter(code %in% low_inc_codes) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "Low income", Country.Code = "LIC")
### fcs average
fcs_avg_s <- country_pops |>
  filter(year == "2021") |>
  right_join(insecurity_s, by = c("code" = "Country.Code")) |>
  filter(code %in% fcs_codes) |>
  summarize(pct = weighted.mean(pct, population, na.rm = TRUE)) |>
  mutate(Country.Name = "FCS Countries", Country.Code = "FCS")
### Joining to main dataset and filtering
insecurity_s <- insecurity_s |>
  filter(Country.Code == "SOM") |>
  bind_rows( wld_avg_s, ssa_avg_s, lic_avg_s, fcs_avg_s)
```


```{r, echo = FALSE}
modsev_plot <- insecurity_ms |>
  mutate(pct = pct/100,
         issomalia = !is.na(Country.Code) & Country.Code == "SOM",
         Country.Name = fct_reorder(Country.Name, -(pct))) |>
  ggplot(aes(x = Country.Name, y = pct, fill = issomalia)) +
  geom_col() +
  theme_minimal() +
  labs(
    title = "2021: Prevalence of Moderate or Severe Food Insecurity, % of Population",
    x = "",
    y = "",
    fill = "Country/Grouping") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#8A969F", "#0071BC")) +
  theme(legend.position = "none")
# ggsave("Visualizations/Mod_Sev_Insecurity.png")
# modsev_plot

sev_plot <- insecurity_s |>
  mutate(pct = pct/100,
         issomalia = !is.na(Country.Code) & Country.Code == "SOM",
         Country.Name = fct_reorder(Country.Name, -(pct))) |>
  ggplot(aes(x = Country.Name, y = pct, fill = issomalia)) +
  theme_minimal() +
  geom_col() +
  labs(
    title = "2021: Prevalence of Severe Food Insecurity, % of Population",
    x = "",
    y = "",
    fill = "Country/Grouping") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#8A969F", "#0071BC")) +
  theme(legend.position = "none")

# ggsave("Visualizations/Sev_Insecurity.png")
# sev_plot

## Stacked Bar Charts
combined_fies <- insecurity_ms |>
  rename("Moderate Only" = pct) |>
  full_join(insecurity_s, by = c("Country.Name", "Country.Code")) |>
  rename("Severe" = pct) |>
  mutate("Moderate Only" = (`Moderate Only` - Severe) / 100,
         "Severe" = Severe / 100) |>
  pivot_longer(cols = 3:4,
               names_to = "level",
               values_to = "value") |>
  mutate(Country.Name = fct_reorder(Country.Name, -(value)))
```



```{r, echo = FALSE, warning = FALSE}
combined_fies_plot_grouped <- insecurity_ms |>
  rename("Moderate and Severe" = pct) |>
  full_join(insecurity_s, by = c("Country.Name", "Country.Code")) |>
  rename("Severe" = pct) |>
  pivot_longer(cols = 3:4,
               names_to = "level",
               values_to = "value") |>
  mutate(value = value / 100,
         Country.Name = fct_reorder(Country.Name, -(value))) |>
  ggplot(aes(x = Country.Name, y = value, fill = level)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "2021: Prevalence of Moderate and Severe Food Insecurity, % of Population",
    x = "",
    y = "",
    fill = "Insecurity Level") + 
  scale_fill_manual(values = c(
  "Moderate and Severe" = "#f89d7b",
  "Severe" = "#bb301e"))
combined_fies_plot_grouped
```

Prevalence of moderate-and-severe and severe food insecurity in Somalia compared to peer country groups, 2021.

```{r, echo = FALSE, warning = FALSE}

combined_fies_plot_stacked <- combined_fies |>
  ggplot(aes(x = Country.Name, y = value, fill = level)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "2021: Prevalence of Moderate and Severe Food Insecurity, % of Population",
    x = "",
    y = "",
    fill = "Insecurity Level") + 
  scale_fill_manual(values = c(
  "Moderate Only" = "#e56245",
  "Severe" = "#bb301e"))
combined_fies_plot_stacked

# '#ffd6b9', '#fdb999', '#f89d7b', '#ef805f', '#e56245', '#d24830', '#bb301e', '#a2180f', '#870000'
## Grouped Bar Charts
```

The same data shown as stacked bars, separating the severe and moderate-only components of food insecurity.


```{r, warning = FALSE, echo = FALSE}
# FCS Comparison charts

fcs_insecurity <- wbstats::wb_data(c("mod_sev" = "SN.ITK.MSFI.ZS", "sev" = "SN.ITK.SVFI.ZS"), country = "countries_only") |>
  filter(iso3c %in% fcs_codes) |>
  filter(!is.na(mod_sev)) |>
  group_by(country) |>
  filter(date == 2021) |>
  ungroup() |>
  mutate(highlight = (iso3c == "SOM"))
  
fcs_ms_plot <- fcs_insecurity |>
  mutate(country = fct_reorder(country, -mod_sev),
         mod_sev = mod_sev / 100) |>
  ggplot(aes(x = country, y = mod_sev, fill = highlight)) + geom_bar(stat = "identity") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "2021: Prevalence of Moderate and Severe Food Insecurity",
    subtitle = "% of Population, 2026 FCS Countries",
    y = "",
    x = ""
  ) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#8A969F", "#0071BC"))
fcs_ms_plot
```

Combined moderate and severe food insecurity rates across all 2026 FCS countries, 2021. Somalia ranks among the highest.

```{r, echo = FALSE, warning = FALSE}
fcs_s_plot <- fcs_insecurity |>
  filter(!is.na(sev)) |>
  mutate(country = fct_reorder(country, -sev),
         sev = sev / 100) |>
  ggplot(aes(x = country, y = sev, fill = highlight)) + geom_bar(stat = "identity") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "2021: Prevalence of Severe Food Insecurity",
    subtitle = "% of Population, 2026 FCS Countries",
    y = "",
    x = ""
  ) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#8A969F", "#0071BC"))
fcs_s_plot
```

Severe food insecurity rates across FCS countries, 2021. Somalia falls in the worst tier alongside Haiti and DRC.


---

## IPC data

```{r  Loading in and Cleaning Data, echo = FALSE}
ipc0 <- read_excel("Data/IPC_Population_Analysis_Styled_2017,2027_1781118839874.xlsx") |>
  select(1:"Current - Phase 3+ %") |>
  filter(!is.na(`Country Population`) & !is.na(`Current - Analysis Period`)) |>
  filter(`Country/Analysis Name/Group Name` != "SO: Acute Food Insecurity Analysis January 2026")
ipc1 <- read_excel("Data/IPC_Population_Analysis_Styled_2017,2027_1781119874050.xlsx") |>
  select(1:"Current - Phase 3+ %") |>
  filter(!is.na(`Country Population`))
ipc <- bind_rows(ipc0, ipc1)
ipc <- ipc |> 
  rename(population = `Country Population`, date = `Date of Analysis`) |>
  mutate(date = ymd_hms(date))

## Data with population
ipc_pop <- ipc |>
  select(population, date, ends_with("Pop")) |>
  rename_with(~gsub("Current - ", "", .x)) |>
  rename_with(~gsub(" Pop", "", .x))
ipc_pop <- ipc_pop |>
  pivot_longer(cols = `Phase 1`:`Phase 3+`,
               names_to = "phase",
               values_to = "pop") |>
  mutate(pop_scaled = pop/1000000)

## Data with percents
ipc_pct <- ipc |>
  select(population, date, ends_with("%")) |>
  rename_with(~gsub("Current - ", "", .x)) |>
  rename_with(~gsub(" %", "", .x))
ipc_pct <- ipc_pct |>
  pivot_longer(cols = `Phase 1`:`Phase 3+`,
               names_to = "phase",
               values_to = "pct")
```



```{r IPC Visualizations, echo = FALSE}
ipc_pop_plot <- ipc_pop |> filter(phase != "Phase 3+") |>
  ggplot(aes(x = date, y = pop_scaled, fill = phase)) +
  theme_minimal() +
  geom_area() +
  labs(
    title = "IPC Food Insecurity Classifications, by Population (Millions)",
    subtitle = "Somalia 2017-2026",
    x = "",
    y = "",
    fill = "") +
  scale_fill_manual(values = c('#e3f6fd', '#b6d4f1', '#8b6486', '#7f3f4d', '#691b15'))
# ggsave("Visualizations/IPC_Phases_Pop.png")
ipc_pop_plot
```

Population in each IPC food insecurity phase in Somalia, 2017–2026, measured in millions of people.

```{r, echo = FALSE, warning = FALSE}

ipc_pct_plot <- ipc_pct |> filter(phase != "Phase 3+") |>
  ggplot(aes(x = date, y = pct, fill = phase)) +
  theme_minimal() +
    geom_area() +
  labs(
    title = "More Recently, Food Security has been Worsening in Somalia",
    subtitle = "IPC Food Insecurity Classifications from 2017-2026, % of Population",
    x = "",
    y = "",
    fill = "") +
  theme(plot.subtitle = element_text(face = "italic")) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c('#e3f6fd', '#b6d4f1', '#8b6486', '#7f3f4d', '#691b15'))
# ggsave("Visualizations/IPC_Phases_Pct.png", width = 7, height = 5)
ipc_pct_plot


# ipc_p3_plot <- ipc_pct |> filter(phase == "Phase 3+") |>
#   ggplot(aes(x = date, y = pct)) + geom_line() +
#   labs(
#     title = ""
#   )
```

The same IPC phase breakdown expressed as a share of Somalia's total assessed population.


```{r, echo = FALSE, warning = FALSE}
ipc_dashboard <- read.csv("Data/IPC_CH_numbers_by_country_29-6-2026.csv", skip = 1) |>
  filter(Coverage != "") |>
  select(-c(Period.from, Period.to, IPC.CH)) |>
  pivot_longer(cols=Phase.1:Phase.5,
               names_to = "Phase",
               values_to = "value") |>
  mutate(highlight = (Country == "Somalia")) |>
  filter(Phase != "Phase.3.") |> 
  mutate(Country = case_when(
    Country == "Democratic Republic of the Congo" ~ "Congo DR",
    Country == "United Republic of Tanzania" ~ "Tanzania",
    TRUE ~ Country)) |>
  mutate(Country = fct_reorder(Country, Phase.3.), 
         Country = fct_reorder(Country, highlight))

ipc_dashboard_plot <- ipc_dashboard |>
  mutate(value = value / 100) |>
  ggplot(aes(x = Country, y = value, fill = Phase)) +
  geom_col() +
  theme_minimal() +
  labs(
    title = "Somalia is Currently Among the Worst IPC Countries",
    subtitle = "Most Recent 2026 IPC/CH Classification by Country, % of Population",
    x = "",
    y = "",
    fill = "") +
  coord_flip() +
  scale_fill_manual(
  values = c('#e3f6fd', '#b6d4f1', '#8b6486', '#7f3f4d', '#691b15'),
  labels = c("Phase.1" = "Phase 1", "Phase.2" = "Phase 2", "Phase.3" = "Phase 3", "Phase.4" = "Phase 4", "Phase.5" = "Phase 5")) +
  scale_y_continuous(breaks = seq(0, 1, by = .20), labels = scales::percent)

ipc_dashboard_plot
```

Most recent IPC/CH classification by country, showing the share of population in each phase. Somalia ranks among the most severe globally.


---

## Stunting 
```{r Loading in Data, message=FALSE, echo = FALSE, warning = FALSE}
stunting_modeled <- wbstats::wb_data(indicator = "SH.STA.STNT.ME.ZS",
                                     start_date = 2000,
                                     end_date = 2026,
                                     lang = "en",
                                     country = benchmarks)


stunting <- wbstats::wb_data(indicator = "SH.STA.STNT.ME.ZS", country = "countries_only")
stunt_progress_results <- suppressWarnings(track_progress(
  data = stunting,
  indicator = "SH.STA.STNT.ME.ZS",
  code_col = "iso3c",
  year_col = "date",
  startyear_data = 2000,
  endyear_data = 2024,
  eval_from = 2015,
  eval_to = 2023,
  speed = TRUE,
  percentiles = FALSE,
  future =  FALSE,
  target_year = 2030,
  sequence_pctl = seq(20, 80, 20),
  sequence_speed = c(0.25, 0.5, 1, 2, 4),
  best = "low",
  support = 1,
  granularity = 0.01
))
```

```{r Create Viz, echo=FALSE, message=FALSE, warning=FALSE}
stunting_modeled_plot <- stunting_modeled |>
  filter(date <= max(date[!is.na(SH.STA.STNT.ME.ZS)])) |>
  mutate(label = if_else(date == max(date), as.character(country), NA_character_),
         highlight = (iso3c == "SOM")) |>
  ggplot(aes(x=date, y = SH.STA.STNT.ME.ZS, group = country, color = highlight)) +
  theme_minimal() +
  geom_line(linewidth = 0.75) +
  labs(
    title = "Prevalence of Stunting, (% of Population)",
    x = "",
    y = "",
    color = "Country/Grouping") +
    annotate("text",
           x = 2021, 
           y = 46,
           label = "2024: Data ends",
           fontface = "italic",
           size = 3.5) +
  geom_vline(xintercept = 2024, linetype = "dashed", color = "black") +
  theme(
    legend.position = "none") +
  scale_color_manual(values = c("#8A969F", "#0071BC")) +
  geom_label_repel(aes(label = label),
                  nudge_x = 7.5,
                  direction = "y",
                  hjust = 0.5,
                  segment.color = NA,
                  na.rm = TRUE)
  
# ggsave("Visualizations/Modeled_Stunting.png")
stunting_modeled_plot
```

Somalia's stunting prevalence compared to regional and income-group averages, 2000–2024.

```{r, echo = FALSE, warning = FALSE}
#Line Chart with FCS Countries
stunting_modeled <- wbstats::wb_data(indicator = "SH.STA.STNT.ME.ZS",
                                     start_date = 2000,
                                     end_date = 2026,
                                     lang = "en",
                                     country = fcs_codes)

stunting_modeled_plot_fcs <- stunting_modeled |>
  filter(date <= max(date[!is.na(SH.STA.STNT.ME.ZS)])) |>
  mutate(label = if_else(date == max(date) & (iso3c == "SOM"), as.character(country), NA_character_),
         highlight = (iso3c == "SOM")) |>
  ggplot(aes(x=date, y = SH.STA.STNT.ME.ZS, group = country, color = highlight)) +
  theme_minimal() +
  geom_line(linewidth = 0.75) +
  labs(
    title = "Prevalence of Stunting, (% of Population)",
    x = "",
    y = "",
    color = "Country/Grouping") +
    annotate("text",
           x = 2021, 
           y = 61.5,
           label = "2024: Data ends",
           fontface = "italic",
           size = 3.5) +
  geom_vline(xintercept = 2024, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("#8A969F", "#0071BC"),
                     labels = c("TRUE" = "Somalia", "FALSE" = "FCS Countries"))
  

# ggsave("Visualizations/Modeled_Stunting.png")
stunting_modeled_plot_fcs

```

Somalia's stunting rate against individual FCS country peers, 2000–2024.

```{r, echo = FALSE, warning = FALSE}

stunting_arrow <- stunting_modeled |>
  select(country, date, SH.STA.STNT.ME.ZS) |>
  filter(date == 2024 | date == 2015) |>
  mutate(SH.STA.STNT.ME.ZS = SH.STA.STNT.ME.ZS / 100) |>
  pivot_wider(names_from = date, values_from = SH.STA.STNT.ME.ZS, names_prefix = "stunt_") |>
  mutate(highlight = (country == "Somalia, Fed. Rep."))

stunting_arrow_plot1 <- stunting_arrow |>
  ggplot() +
  geom_segment(aes(x = reorder(country, -stunt_2024), xend = reorder(country, stunt_2024),
                   y = stunt_2015, yend = stunt_2024, color = highlight),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed")) +
  geom_point(aes(x = reorder(country, -stunt_2024), y = stunt_2015, color = highlight), size = 1) +
  scale_color_manual(
    values = c("TRUE" = "#0071BC", "FALSE" = "#8A969F"),
    labels = c("TRUE" = "Somalia", "FALSE" = "FCS Countries"),
    name = NULL
  ) +
  labs(x = "",
       y = "",
       title = "Prevalence of Stunting Change from 2015-2024",
       subtitle = "% of Population, Sorted by End Year Value") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  coord_flip()
stunting_arrow_plot1


stunting_arrow_plot2 <- stunting_arrow |>
  ggplot() +
  geom_segment(aes(x = reorder(country, -stunt_2015), xend = reorder(country, stunt_2015),
                   y = stunt_2015, yend = stunt_2024,
                   color = highlight),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed")) +
  geom_point(aes(x = reorder(country, -stunt_2015), y = stunt_2015, color = highlight), size = 1) +
  scale_color_manual(
    values = c("TRUE" = "#0071BC", "FALSE" = "#8A969F"),
    labels = c("TRUE" = "Somalia", "FALSE" = "FCS Countries"),
    name = NULL
  ) +
  labs(x = "",
       y = "",
       title = "Prevalence of Stunting Change from 2015-2024",
       subtitle = "% of Population, Sorted by Start Year Value") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  coord_flip()
stunting_arrow_plot2

```

Change in stunting prevalence across FCS countries from 2015 to 2024, ranked by 2024 rate and 2015 rate. Somalia falls in the middle of the group.




---

## Speed of Progress
### Undernourishment
```{r Getting Data, message = FALSE, echo = FALSE}
data <- wbstats::wb_data(indicator = "SN.ITK.DEFC.ZS", lang = "en", country = "countries_only")
progress_results <- suppressWarnings(track_progress(data = data,
                                   indicator = "SN.ITK.DEFC.ZS",
                                   code_col = "iso3c",
                                   year_col = "date",
                                   startyear_data = 2001,
                                   endyear_data = 2023,
                                   eval_from = 2015,
                                   eval_to = 2023, 
                                   speed = TRUE,
                                   percentiles = FALSE,
                                   future = FALSE, ## TRUE,
                                   target_year = 2030,
                                   sequence_pctl = seq(20,80,20),
                                   sequence_speed = c(0.25, 0.5, 1, 2, 4),
                                   best = "low",
                                   support = 1,
                                   granularity = 0.01))


## Cleaning
score_speed <- classifications |>
  select(Economy, Code) |>
  right_join(progress_results$scores$speed, by = c("Code" = "code")) |>
  rename("country" = Economy, "code" = Code)

score_speed_fcs_separate <- score_speed |>
  filter(code %in% fcs_codes) |>
  mutate(country = fct_reorder(country, score)) |>
  mutate(issomalia = (code == "SOM"))
```


```{r Creating plot, echo = FALSE}
fcs_pou_plot <- score_speed_fcs_separate |>
  ggplot(aes(x = country, y = score, fill = issomalia)) + geom_bar(stat="identity") +
  theme_minimal() +
  coord_flip() +
  labs(
    title = "Speed of Progress Measured from 2015-2023",
    subtitle = "2026 FCV Countries",
    y = "Speed of Progress",
    x = ""
  ) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#8A969F", "#0071BC")) +
  theme(plot.subtitle = element_text(face = "italic"))
fcs_pou_plot
ggsave("Visualizations/FCV_PoU.png", width = 7, height = 5)
```

Speed of Progress of undernourishment for Somalia and other FCS Countries, 2015–2023. Speed of progress is calculated by taking historical data, generating a "typical development" path, and comparing the rate of improvement from a given start point. Higher values indicate faster improvement. Despite have a speed score of < 1, it is clear that Somalia is faring better than most FCS Countries.

### Stunting
```{r Creating plot for stunt, echo = FALSE}
fcs_stunt_plot <- stunt_progress_results$scores$speed |>
  filter(code %in% fcs_codes) |>
  mutate(highlight = (code == "SOM")) |>
  left_join(classifications, by = c("code" = "Code")) |>
  mutate(Economy = fct_reorder(Economy, score)) |>
  ggplot(aes(x = Economy, y = score, fill = highlight)) + 
  theme_minimal() +
  geom_col() +
  coord_flip() +
  labs(
    title = "Speed of Progress from 2015-2023: Stunting",
    subtitle = "2026 FCS Countries",
    y = "Speed of Progress",
    x = ""
  ) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#8A969F", "#0071BC"))
fcs_stunt_plot
```

The same graph as above but with stunting.


---

## Typical Path for Undernourishment and Stunting with Entire World
### Undernourishment
``` {r World Undernourishment, message = FALSE, echo = FALSE, warning = FALSE}

## running on undernourishment track progress
data <- wbstats::wb_data(indicator = "SN.ITK.DEFC.ZS", lang = "en", country = "countries_only")
progress_results <- suppressWarnings(track_progress(data = data,
                                   indicator = "SN.ITK.DEFC.ZS",
                                   code_col = "iso3c",
                                   year_col = "date",
                                   startyear_data = 2001,
                                   endyear_data = 2023,
                                   eval_from = 2015,
                                   eval_to = 2023, 
                                   speed = TRUE,
                                   percentiles = FALSE,
                                   future = FALSE, ## TRUE,
                                   target_year = 2030,
                                   sequence_pctl = seq(20,80,20),
                                   sequence_speed = c(0.25, 0.5, 1, 2, 4),
                                   best = "low",
                                   support = 1,
                                   granularity = 0.01))

check <- progress_results$scores$speed
path_speed  <- progress_results$predicted_changes$path_speed

## Combining and Plotting Somalia's path and the typical path
pou_som <- pou |>
  mutate(year = as.numeric(year)) |>
  filter(Country.Code == "SOM",
         year >= 2015) |>
  mutate(time = as.numeric(year) - 2015) |>
  arrange(time)
pou_som_yr_offset <- as.numeric(path_speed[which.min(abs(path_speed$y-pou_som$value[1])), "time"])
pou_som <- pou_som |>
  mutate(time = time + pou_som_yr_offset, 
         label = if_else(year == max(year) | year == min(year), as.character(year), NA_character_))

combined_path_speed <- path_speed |>
  mutate(Country.Name = "Typical Path") |>
  rename("value" = y) |>
  full_join(pou_som, by = c("Country.Name", "time", "value"))


combined_path_speed_plot <- combined_path_speed |>
  ggplot(aes(x = time, y = value, color = Country.Name)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  labs(
    title = "Worldwide Typical Development Path",
    subtitle = "Prevalence of Undernourishment, % of Population",
    x = "Years of Progress",
    y = "",
    color = "") +
  scale_color_manual(values = c("#0071BC", "#111111")) +
  geom_line(data = combined_path_speed |> filter(Country.Name != "Typical Path"), linewidth = 1.25, color = "#0071BC") +
  geom_label_repel(aes(label = label),
                  nudge_x = 6,
                  show.legend = FALSE)

combined_path_speed_plot
```

Somalia's progress from 2015-2023 in undernourishment rates positioned on the typical development path, which maps how countries have historically reduced undernourishment as they develop.

```{r, echo = FALSE, warning = FALSE}
## FCS Countries

fcs_pou <- wbstats::wb_data(c( "pou" = "SN.ITK.DEFC.ZS"), country = fcs_codes) |>
  select(country, date, pou) |>
  filter(date >= 2015 & !is.na(pou))

offsets <- fcs_pou |>
  group_by(country) |>
  slice_min(date, n = 1) |>  # first observation per country
  ungroup() |>
  rowwise() |>
  mutate(yr_offset = as.numeric(path_speed[which.min(abs(path_speed$y - pou)), "time"])) |>
  ungroup() |>
  select(country, yr_offset)

fcs_pou <- fcs_pou |>
  left_join(offsets, by = "country") |>
  mutate(time = (date - 2015) + yr_offset) |>
  select(country, time, pou)

fcs_wld_pou <- path_speed |>
  mutate(country = "Typical Path") |>
  rename("pou" = y) |>
  full_join(fcs_pou, by = c("country", "time", "pou"))


typical_path <- fcs_wld_pou |> filter(country == "Typical Path")
fcs_countries <- fcs_wld_pou |> filter(!country %in% c("Typical Path", "Somalia"))
somalia <- fcs_wld_pou |> filter(country == "Somalia, Fed. Rep.")

fcs_wld_pou_plot <- ggplot() +
  geom_line(data = typical_path, aes(x = time, y = pou), 
            color = "#111111", linewidth = 0.9) +
  geom_line(data = fcs_countries, aes(x = time, y = pou, group = country), 
            color = "#8A969F", linewidth = 0.5) +
  geom_line(data = somalia, aes(x = time, y = pou), 
            color = "#0071BC", linewidth = 1.25) +
  # dummy data for legend
  geom_line(data = data.frame(time = NA, pou = NA, label = factor(c("Typical Path", "FCS Countries", "Somalia, Fed. Rep."))),
            aes(x = time, y = pou, color = label)) +
  scale_color_manual(values = c("Typical Path" = "#111111", "FCS Countries" = "#8A969F", "Somalia, Fed. Rep." = "#0071BC")) +
  theme_minimal() +
  labs(
    title = "For FCV Countries, Being Average is Above Average",
    subtitle = "Typical Development Path for Prevalence of Undernourishment, as a % of Population",
    x = "Years of Progress",
    y = "",
    color = "") +
  theme(plot.subtitle = element_text(face = "italic"))
fcs_wld_pou_plot
ggsave("Visualizations/FCV_SoP_PoU.png", width = 7, height = 5)




```

FCS country trajectories overlaid on the typical development path for undernourishment, 2015–2023. Many of the countries trend upward, signalling a regression in this indicator.

### Stunting
``` {r World Stunting, message = FALSE, echo = FALSE, warning = FALSE}

## running on stunting track progress
data <- stunting
progress_results <- suppressWarnings(track_progress(data = data,
                                   indicator = "SH.STA.STNT.ME.ZS",
                                   code_col = "iso3c",
                                   year_col = "date",
                                   startyear_data = 1960,
                                   endyear_data = 2025,
                                   eval_from = 2015,
                                   eval_to = 2023, 
                                   speed = TRUE,
                                   percentiles = FALSE,
                                   future = FALSE, ## TRUE,
                                   target_year = 2030,
                                   sequence_pctl = seq(20,80,20),
                                   sequence_speed = c(0.25, 0.5, 1, 2, 4),
                                   best = "low",
                                   support = 1,
                                   granularity = 0.01))


check <- progress_results$scores$speed
path_speed  <- progress_results$predicted_changes$path_speed

## Combining and Plotting Somalia's path and the typical path
path_speed  <- progress_results$predicted_changes$path_speed

stunt_som <- stunting |>
  filter(iso3c == "SOM", 
         date >= 2015,
         !is.na(SH.STA.STNT.ME.ZS)) |>
  mutate(time = date - 2015) |>
  arrange(time) |>
  select(country, SH.STA.STNT.ME.ZS, time, date)

stunt_som_yr_offset <- as.numeric(path_speed[which.min(abs(path_speed$y-stunt_som$SH.STA.STNT.ME.ZS[1])), "time"])

stunt_som <- stunt_som |>
  mutate(time = time + stunt_som_yr_offset,
         label = if_else(date == max(date) | date == min(date), as.character(date), NA_character_))

combined_path_speed <- path_speed |>
  mutate(country = "Typical Path") |>
  rename("SH.STA.STNT.ME.ZS" = y) |>
  full_join(stunt_som, by = c("country", "time", "SH.STA.STNT.ME.ZS"))
  
combined_path_speed_plot <- combined_path_speed |>
  ggplot(aes(x = time, y = SH.STA.STNT.ME.ZS, color = country)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  labs(
    title = "Worldwide Typical Development Path",
    subtitle = "Prevalence of stunting, height for age (modeled estimate, % of children under 5)",
    x = "Years of Progress",
    y = "",
    color = "") +
  scale_color_manual(values = c("#0071BC", "#111111")) +
  geom_line(data = combined_path_speed |> filter(country != "Typical Path"), linewidth = 1.25, color = "#0071BC") +
  geom_label_repel(aes(label = label),
                  nudge_x = 6,
                  show.legend = FALSE)
  
combined_path_speed_plot
```

Somalia's 2015-2024 stunting rate progress on the global typical development path.

```{r, echo = FALSE, warning = FALSE}
## FCS Countries

fcs_stunt <- wbstats::wb_data(c( "stunt" = "SH.STA.STNT.ME.ZS"), country = fcs_codes) |>
  select(country, date, stunt) |>
  filter(date >= 2015 & !is.na(stunt))

offsets <- fcs_stunt |>
  group_by(country) |>
  slice_min(date, n = 1) |>  # first observation per country
  ungroup() |>
  rowwise() |>
  mutate(yr_offset = as.numeric(path_speed[which.min(abs(path_speed$y - stunt)), "time"])) |>
  ungroup() |>
  select(country, yr_offset)

fcs_stunt <- fcs_stunt |>
  left_join(offsets, by = "country") |>
  mutate(time = (date - 2015) + yr_offset) |>
  select(country, time, stunt)

fcs_wld_stunt <- path_speed |>
  mutate(country = "Typical Path") |>
  rename("stunt" = y) |>
  full_join(fcs_stunt, by = c("country", "time", "stunt"))


typical_path <- fcs_wld_stunt |> filter(country == "Typical Path")
fcs_countries <- fcs_wld_stunt |> filter(!country %in% c("Typical Path", "Somalia"))
somalia <- fcs_wld_stunt |> filter(country == "Somalia, Fed. Rep.")

fcs_wld_stunt_plot <- ggplot() +
  geom_line(data = typical_path, aes(x = time, y = stunt), 
            color = "#111111", linewidth = 0.9) +
  geom_line(data = fcs_countries, aes(x = time, y = stunt, group = country), 
            color = "#8A969F", linewidth = 0.5) +
  geom_line(data = somalia, aes(x = time, y = stunt), 
            color = "#0071BC", linewidth = 1.25) +
  # dummy data for legend
  geom_line(data = data.frame(time = NA, stunt = NA, label = factor(c("Typical Path", "FCS Countries", "Somalia, Fed. Rep."))),
            aes(x = time, y = stunt, color = label)) +
  scale_color_manual(values = c("Typical Path" = "#111111", "FCS Countries" = "#8A969F", "Somalia, Fed. Rep." = "#0071BC")) +
  theme_minimal() +
  labs(
    title = "FCS Country Development Paths 2015-2023",
    subtitle = "Prevalence of Undernourishment, % of Population",
    x = "Years of Progress",
    y = "",
    color = ""
  )
fcs_wld_stunt_plot

```

FCS country trajectories on the typical stunting development path, 2015–2024.

---


```{r FCS Undernourishment, message = FALSE, include = FALSE}

## running on undernourishment track progress
data <- wbstats::wb_data(indicator = "SN.ITK.DEFC.ZS", lang = "en", country = "countries_only") |>
  filter(iso3c %in% fcs_codes)
progress_results <- suppressWarnings(track_progress(data = data,
                                   indicator = "SN.ITK.DEFC.ZS",
                                   code_col = "iso3c",
                                   year_col = "date",
                                   startyear_data = 2001,
                                   endyear_data = 2023,
                                   eval_from = 2015,
                                   eval_to = 2023, 
                                   speed = TRUE,
                                   percentiles = FALSE,
                                   future = FALSE, ## TRUE,
                                   target_year = 2030,
                                   sequence_pctl = seq(20,80,20),
                                   sequence_speed = c(0.25, 0.5, 1, 2, 4),
                                   best = "low",
                                   support = 1,
                                   granularity = 0.01))

check <- progress_results$scores$speed
path_speed  <- progress_results$predicted_changes$path_speed

## Combining and Plotting Somalia's path and the typical path
pou_som <- pou |>
  mutate(year = as.numeric(year)) |>
  filter(Country.Code == "SOM",
         year >= 2015) |>
  mutate(time = as.numeric(year) - 2015) |>
  arrange(time)
pou_som_yr_offset <- as.numeric(path_speed[which.min(abs(path_speed$y-pou_som$value[1])), "time"])
pou_som <- pou_som |>
  mutate(time = time + pou_som_yr_offset)

combined_path_speed <- path_speed |>
  mutate(Country.Name = "Typical Path") |>
  rename("value" = y) |>
  full_join(pou_som, by = c("Country.Name", "time", "value"))
  
combined_path_speed_plot <- combined_path_speed |>
  ggplot(aes(x = time, y = value, color = Country.Name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "FCS Typical Development Path",
    subtitle = "Prevalence of Undernourishment, % of Population",
    x = "Years",
    y = "",
    color = "") +
  scale_color_manual(values = c("#0071BC", "#8A969F"))
  
# combined_path_speed_plot
```

``` {r FCS Stunting, message = FALSE, include = FALSE}

## running on stunting track progress
data <- stunting |>
  filter(iso3c %in% fcs_codes)
progress_results <- suppressWarnings(track_progress(data = data,
                                   indicator = "SH.STA.STNT.ME.ZS",
                                   code_col = "iso3c",
                                   year_col = "date",
                                   startyear_data = 1960,
                                   endyear_data = 2025,
                                   eval_from = 2015,
                                   eval_to = 2023, 
                                   speed = TRUE,
                                   percentiles = FALSE,
                                   future = FALSE, ## TRUE,
                                   target_year = 2030,
                                   sequence_pctl = seq(20,80,20),
                                   sequence_speed = c(0.25, 0.5, 1, 2, 4),
                                   best = "low",
                                   support = 1,
                                   granularity = 0.01))


check <- progress_results$scores$speed
path_speed  <- progress_results$predicted_changes$path_speed

## Combining and Plotting Somalia's path and the typical path
path_speed  <- progress_results$predicted_changes$path_speed

stunt_som <- stunting |>
  filter(iso3c == "SOM", 
         date >= 2015) |>
  mutate(time = date - 2015) |>
  arrange(time) |>
  select(country, SH.STA.STNT.ME.ZS, time)

stunt_som_yr_offset <- as.numeric(path_speed[which.min(abs(path_speed$y-stunt_som$SH.STA.STNT.ME.ZS[1])), "time"])

stunt_som <- stunt_som |>
  mutate(time = time + stunt_som_yr_offset)

combined_path_speed <- path_speed |>
  mutate(country = "Typical Path") |>
  rename("SH.STA.STNT.ME.ZS" = y) |>
  full_join(stunt_som, by = c("country", "time", "SH.STA.STNT.ME.ZS"))
  
combined_path_speed_plot <- combined_path_speed |>
  ggplot(aes(x = time, y = SH.STA.STNT.ME.ZS, color = country)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "FCS Typical Development Path",
    subtitle = "Prevalence of stunting, height for age (modeled estimate, % of children under 5)",
    x = "Years",
    y = "",
    color = "") +
  scale_color_manual(values = c("#0071BC", "#8A969F"))
  
# combined_path_speed_plot

```











```{r testing things, include = FALSE}
somalia_world_hunger <- wbstats::wb_data(indicator = c(pop = "SP.POP.TOTL", mod_sev = "SN.ITK.MSFI.ZS"),
                                         country = c("SOM", "WLD"))

somalia_world_hunger <- somalia_world_hunger |>
  filter(date == "2021") |>
  mutate(num_hunger = (mod_sev / 100) * pop)

s






```


















